---
title: "Endotherm Model Tutorial I: component functions"
author: "Michael Kearney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Endotherm Components Tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 eval = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE  
)
```

## Overview

This document provides an overview of an R implementation of the NicheMapR endotherm model, via the sub-functions of 'endoR'. This implementation modularises the original version such that many of the internal subroutines and functions can be called separately. The idea is that the user has more flexibility in setting the model up for the behavioural and physiological responses of a particular animal. There is also a tutorial for using 'endoR' itself [Endotherm Model Tutorial](endotherm-model-tutorial.html).

```{r}
library(NicheMapR)
library(knitr)
library(plotrix)
```

## Running the separate subroutines

### IRPROP

This subroutine computes the feasible hair spacing and the three parameters needed for computing conduction and infrared radiation through the fur. It assumes a thermal conductivity of hair (keratin) of 0.209 W m{-1} K{-1} and that the thermal conductivity of air is 0.02425+(7.038e-5*TA).

```{r}
# environmental input
TA <- 20 # air temperature, for calculation of conductivity of air (°C)

# shape input
SHAPE_B_MAX <- 2.7 # max possible ratio between long and short axis (-)
SHAPE_B_REF <- 2.7 # initial ratio between long and short axis (-)
SHAPE_B <- 2.7 # current ratio between long and short axis (-)
MAXPTVEN <- 0.3 # maxium fraction of surface area that is ventral (fractional, 0-1)

# fur properties
DHAIRD <- 30E-06 # hair diameter, dorsal (m)
DHAIRV <- 30E-06 # hair diameter, ventral (m)
LHAIRD <- 23.9E-03 # hair length, dorsal (m)  
LHAIRV <- 23.9E-03 # hair length, ventral (m)  
ZFURD <- 9E-03 # fur depth, dorsal (m)
ZFURV <- 9E-03 # fur depth, ventral (m)
ZFURCOMP <- ZFURV # depth of compressed fur (for conduction) (m)
RHOD <- 3968E+04 # hair density, dorsal (1/m2) 
RHOV <- 2781E+04 # hair density, ventral (1/m2)
REFLD <- 0.301  # fur reflectivity dorsal (fractional, 0-1) 
REFLV <- 0.301  # fur reflectivity ventral (fractional, 0-1)

# call the subroutine
IRPROP.out <- IRPROP(TA, SHAPE_B_MAX, SHAPE_B_REF, SHAPE_B, DHAIRD, DHAIRV, LHAIRD, LHAIRV, ZFURD, ZFURV, RHOD, RHOV, REFLD, REFLV, MAXPTVEN, ZFURCOMP)

# output
E4B1 <- IRPROP.out[1] # exponential integral, E4, of optical thickness, B1 (m), i.e. E4(B1)
KEFARA <- IRPROP.out[2:4] # effective thermal conductivity of fur array, mean, dorsal, ventral (W/mK)
BETARA <- IRPROP.out[5:7] # term involved in computing optical thickess
B1ARA <- IRPROP.out[8:10] # optical thickness array, mean, dorsal, ventral (-)
DHAR <- IRPROP.out[11:13] # fur diameter array, mean, dorsal, ventral (m)
LHAR <- IRPROP.out[14:16] # fur length array, mean, dorsal, ventral (m)
RHOAR <- IRPROP.out[17:19] # fur density array, mean, dorsal, ventral (fibers/m2)  
ZZFUR <- IRPROP.out[20:22] # fur depth array, mean, dorsal, ventral (m)  
REFLFR <- IRPROP.out[23:25] # fur reflectivity array, mean, dorsal, ventral (fractional, 0-1)
FURTST <- IRPROP.out[26] # test of presence of fur (length x diamater x density x depth) (-)
KFURCMPRS <- IRPROP.out[27] # effictive thermal conductivity of compressed ventral fur (W/mK)

IRPROP.lab <- c("E4B1", "KEFARA mean", "KEFARA dorsal", "KEFARA ventral", "BETARA mean", "BETARA dorsal", "BETARA ventral", "B1ARA mean", "B1ARA dorsal", "B1ARA ventral", "DHAR mean", "DHAR dorsal", "DHAR ventral", "LHAR mean", "LHAR dorsal", "LHAR ventral", "RHOAR mean", "RHOAR dorsal", "RHOAR ventral", "ZZFUR mean", "ZZFUR dorsal", "ZZFUR ventral", "REFLFR mean", "REFLFR dorsal", "REFLFR ventral", "FURTST", "KFURCMPRS")
kable(cbind(IRPROP.lab, IRPROP.out[1:27]))
```

Example analysis of fur diameter on effective thermal conductivity of fur.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
DHAIRs <- seq(0, 150, 2) # hair diameters (micrometers)
KEFARAs <- NULL
for(i in 1:length(DHAIRs)){
  KEFARAs[i] <- IRPROP(TA, SHAPE_B_MAX, SHAPE_B_REF, SHAPE_B, DHAIRs[i] * 1E-06, DHAIRs[i] * 1E-06, LHAIRD, LHAIRV, ZFURD, ZFURV, RHOD, RHOV, REFLD, REFLV, MAXPTVEN, ZFURCOMP)[2]
}
plot(KEFARAs ~ DHAIRs, type = 'p', pch = 16, ylab = 'effective fur conductivity, W K-1 m-1', xlab = 'hair diameter, um')
```

Example analysis of fur density on effective thermal conductivity of fur.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
RHOs <- seq(0, 50000, 500) # hair densities (1/cm2)
KEFARAs <- NULL
for(i in 1:length(RHOs)){
  KEFARAs[i] <- IRPROP(TA, SHAPE_B_MAX, SHAPE_B_REF, SHAPE_B, DHAIRD, DHAIRV, LHAIRD, LHAIRV, ZFURD, ZFURV, RHOs[i] * 1E+04, RHOs[i] * 1E+04, REFLD, REFLV, MAXPTVEN, ZFURCOMP)[2]
}
plot(KEFARAs ~ RHOs, type = 'p', pch = 16, ylab = 'effective fur conductivity, W K-1 m-1', xlab = 'hair density, 1/cm2')
```

Example analysis of fur depth on effective thermal conductivity of fur.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
ZFURs <- seq(0, 50, 1) # hair depths (mm)
KEFARAs <- NULL
for(i in 1:length(ZFURs)){
  KEFARAs[i] <- IRPROP(TA, SHAPE_B_MAX, SHAPE_B_REF, SHAPE_B, DHAIRD, DHAIRV, LHAIRD, LHAIRV, ZFURs[i] * 1E-03, ZFURs[i] * 1E-03, RHOD, RHOV, REFLD, REFLV, MAXPTVEN, ZFURCOMP)[2]
}
plot(KEFARAs ~ ZFURs, type = 'p', pch = 16, ylab = 'effective fur conductivity, W K-1 m-1', xlab = 'fur depth, mm')
```

### GEOM

This subroutine computes aspects of the animal's geometry. Note that GEOM requires outputs from IRPROP on fur properties. Note also that this routine is called 'allom' in past versions of 'Niche Mapper'.

```{r}
# input
AMASS <- 10 # kg
ANDENS <- 1000 # kg/m3
SUBQFAT <- 1 # is subcutaneous fat present? (0 is no, 1 is yes)
FATPCT <- 20 # % body fat
SHAPE <- 4 # shape, 1 is cylinder, 2 is sphere, 3 is plate, 4 is ellipsoid
SHAPE_B_REF <- 2.7 # initial ratio between long and short axis (-)
SHAPE_B <- 2.7 # current ratio between long and short axis (-)
SHAPE_C <- SHAPE_B # current ratio of length:height (plate)
DHARA <- DHAR[1] # fur diameter, mean (m) (from IRPROP)
RHOARA <- RHOAR[1] # hair density, mean (1/m2) (from IRPROP)
ZFUR <- ZZFUR[1] # fur depth, mean (m) (from IRPROP)
PCOND <- 0.1 # fraction of body in contact with substrate (fractional, 0-1)
BIRD <- 0 # 
SAMODE <- 0 # if 1, uses bird skin surface area scaling from Walsberg and 1978. JEB Biology 76:185–189, if 2, uses mammal surface area scaling from Stahl (1967) J. of App. Physiology, 453–460.
ORIENT <- 0 # if 1, largest surface area normal to sun's ray's, if 2, largest surface parallel to sun's rays, if 0, average of normal/parallel posture, 

# call the subroutine
GEOM.out <- GEOM(AMASS, ANDENS, FATPCT, SHAPE, ZFUR, SUBQFAT, SHAPE_B, SHAPE_B_REF, SHAPE_C, DHARA, RHOARA, PCOND, SAMODE, ORIENT)

# output
VOL <- GEOM.out[1] # volume, m3
D <- GEOM.out[2] # characteristic dimension for convection, m
MASFAT <- GEOM.out[3] # mass body fat, kg
VOLFAT <- GEOM.out[4] # volume body fat, m3
ALENTH <- GEOM.out[5] # length, m
AWIDTH <- GEOM.out[6] # width, m
AHEIT <- GEOM.out[7] # height, m
ATOT <- GEOM.out[8] # total area at fur/feathers-air interface, m2
ASIL <- GEOM.out[9] # silhouette area to use in solar calcs, m2 may be normal, parallel or average set via ORIENT
ASILN <- GEOM.out[10] # silhouette area normal to sun, m2
ASILP <- GEOM.out[11] # silhouette area parallel to sun, m2
GMASS <- GEOM.out[12] # mass, g
AREASKIN <- GEOM.out[13] # area of skin, m2
FLSHVL <- GEOM.out[14] # flesh volume, m3
FATTHK <- GEOM.out[15] # fat layer thickness, m
ASEMAJ <- GEOM.out[16] # semimajor axis length, m
BSEMIN <- GEOM.out[17] # b semiminor axis length, m
CSEMIN <- GEOM.out[18] # c semiminor axis length, m (currently only prolate spheroid)
CONVSK <- GEOM.out[19] # area of skin for evaporation (total skin area - hair area), m2
CONVAR <- GEOM.out[20] # area for convection (total area minus ventral area, as determined by PCOND), m2
R1 <- GEOM.out[21] # shape-specific core-skin radius in shortest dimension, m
R2 <- GEOM.out[22] # shape-specific core-fur radius in shortest dimension, m

# nest properties
NESTYP <- 3 # for nest calculations - 0 = none, 1 = flat, 2 = cup, 3 = cylinder, 4 = half cylinder, 5 = sphere, 6 = dome
D_NEST <- 0.35 # Outer diameter(m)
TKNEST <- 0.01 # Nest wall thickness (m) 
RONEST <- D_NEST / 2
RINEST <- RONEST - TKNEST
if(RINEST < R2){
 # NEST INNER DIAMETER GREATER THAN ANIMAL WITH OR WITHOUT FUR OUTER DIAMETER. ENLARGE NEST
 RINEST <- R2
 RONEST <- RINEST + TKNEST
}
DENEST <- 1 # Density nest material (kg/m3) 
THCONW <- 0.071 # Nest wall (wood: 0.10-0.35;sheep wool:0.05) thermal conductivity (W/m-C) #!
ABSHEL <- 0.71 # Nest solar absorptivity (decimal: 1.0 = 100%)
EMISHEL <- 0.95 # NEest emissivity
SHELEN <- 0.3 # Length(m) #!

GEOM.lab <- c("VOL", "D", "MASFAT", "VOLFAT", "ALENTH", "AWIDTH", "AHEIT", "ATOT", "ASIL", "ASILN", "ASILP", "GMASS", "AREASKIN", "FLSHVL", "FATTHK", "ASEMAJ", "BSEMIN", "CSEMIN", "CONVSK", "CONVAR", "R1", "R2")
kable(cbind(GEOM.lab, t(GEOM.out)))
```

Plot of current animal configuration.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
if(SHAPE == 4){ #ellipsoid
par(mfrow=c(1,2))
plot(c(0,ASEMAJ*2+ZFUR*2), c(0,ASEMAJ*2+ZFUR*2), type="n", main="ellipsoid, sagittal section", ylab = 'minor axis, m', xlab = 'major axis, m', asp=1)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="black", border = "black", a = ASEMAJ+ZFUR, b = BSEMIN+ZFUR)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="pink", border = "pink", a = ASEMAJ, b = BSEMIN)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="red", border = "red", a = ASEMAJ-FATTHK, b = BSEMIN-FATTHK)

plot(c(0,ASEMAJ*2+ZFUR*2), c(0,ASEMAJ*2+ZFUR*2), type="n", main="ellipsoid, transverse section", ylab = 'minor axis, m', xlab = 'minor axis, m', asp=1)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="black", border = "black", a = BSEMIN+ZFUR, b = CSEMIN+ZFUR)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="pink", border = "pink", a = BSEMIN, b = CSEMIN)
draw.ellipse(ASEMAJ+ZFUR, ASEMAJ+ZFUR, col="red", border = "red", a = BSEMIN-FATTHK, b = CSEMIN-FATTHK)
par(mfrow=c(1,1))
}
```

### F_FACTOR

This subroutine computes animal configuration factors for the sky and overhead vegetation, given the shade level.

```{r}
# environmental inputs
SHADE <- 0 # shade level (%)
NITESHAD <- 0 # flag for if animal is behaviourally seeking shade for warmth at night - remove?
QSOLR <- 1000 # solar radiation, horizontal plane (W/m2)

# trait inputs
EMISAN <- 0.99 # animal emissivity (-)
FATOBJ <- 0 # configuration factor to nearby object
FGDREF <- 0.4 # reference configuration factor to ground
FSKREF <- 0.4 # configuration factor to sky

F_FACTOR.out <- F_FACTOR(SHADE, NITESHAD, QSOLR, FATOBJ, NESTYP, RONEST, R1, FGDREF, FSKREF, AREASKIN, EMISAN)

FAVEG <- F_FACTOR.out[1] # configuration factor to vegetation
FASKY <- F_FACTOR.out[2] # configuration factor to sky
FAGRD <- F_FACTOR.out[3] # configuration factor to ground
FANEST <- F_FACTOR.out[4] # configuration factor to nest wall
# constants for infra-red exchange calculatiosn AREASKIN*CONFIG*EMISAN*SIG
C3 <- F_FACTOR.out[5] # sky
C4 <- F_FACTOR.out[6] # ground
C5 <- F_FACTOR.out[7] # object
C6 <- F_FACTOR.out[8] # vegetation (shade)
C7 <- F_FACTOR.out[9] # nest

F_FACTOR.lab <- c("FAVEG", "FASKY", "FAGRD", "FANEST", "C3", "C4", "C5", "C6", "C7")
kable(cbind(F_FACTOR.lab, t(F_FACTOR.out)))
```

### SOLAR

Compute the solar radiation absorbed by different parts of the organism and its environment.

```{r}
# environmental variables
QSOLR <- QSOLR # solar radiation, horizontal plane (W/m2) (defined above for F_FACTOR)
SHADE <- SHADE # shade (fractional, 0-1) (defined above for F_FACTOR)
Z <- 20 # zenith angle of sun (degrees from overhead)
ABSSB <- 0.8 # solar absorptivity of substrate (fractional, 0-1)

# traits
AREA <- ATOT # surface area for solar exchange, m2 (from GEOM)
ABSAND <- 0.8 # solar absorptivity of dorsal fur (fractional, 0-1)
ABSANV <- 0.8 # solar absorptivity of ventral fur (fractional, 0-1)
ASIL <- ASIL # silhouette area normal to sun, m2 (from GEOM)
PCTDIF <- 0.15 # proportion of solar radiation that is diffuse (fractional, 0-1)
FASKY <- FASKY # configuration factor to sky (-) (from F_FACTOR)
FATOBJ <- FATOBJ # configuration factor to object (-) (from F_FACTOR)
FAVEG <- FAVEG # configuration factor to vegetation (-) (from F_FACTOR)

# solar radiation normal to sun's rays
ZEN <- pi/180*Z # convert degrees to radians
if(Z < 90){ # compute solar radiation on a surface normal to the direct rays of the sun
  CZ = cos(ZEN)
  QNORM = QSOLR/CZ
}else{ # diffuse skylight only
  QNORM = QSOLR
}

SOLAR.out <- SOLAR(AREA, ABSAND, ABSANV, ABSSB, ASIL, PCTDIF, QNORM, SHADE, QSOLR, FASKY, FATOBJ, FAVEG)

QSOLAR <- SOLAR.out[1] # total (global) solar radiation (W)
QSDIR <- SOLAR.out[2] # direct solar radiaton (W)
QSOBJ <- SOLAR.out[3] # lateral diffuse solar radiation (W)
QSSKY <- SOLAR.out[4] # diffuse solar radiation from sky (W)
QSRSB <- SOLAR.out[5] # diffuse solar radiation reflected from substrate (W)
QSDIFF <- SOLAR.out[6] # total diffuse solar radiation (W)
QDORSL <- SOLAR.out[7] # dorsal direct solar radiation (W)
QVENTR <- SOLAR.out[8] # ventral diffuse solar radiaton (W)

SOLAR.lab <- c("QSOLAR", "QSDIR", "QSOBJ", "QSSKY", "QSRSB", "QSDIFF", "QDORSL", "QVENTR")
kable(cbind(SOLAR.lab, t(SOLAR.out)))
```

### CONV

Compute convective heat exchange.

```{r}
# input
TS <- 33 # skin temperature (°C)
TENV <- 20 # air temperature (°C)
TFA <- 10 # fur/air interface temperature (°C)
SHAPE <- 4 # shape, 1 is cylinder, 2 is sphere, 3 is plate, 4 is ellipsoid
SURFAR <- CONVAR # surface area for convection, m2 (from GEOM)
FLTYPE <- 0 # FLUID TYPE: 0 = AIR; 1 = FRESH WATER; 2 = SALT WATER
FURTST <- FURTST # test of fur presence (-) from IRPROP 
VEL <- 1 # wind speed (m/s)
ELEV <- 0 # altitude (m)
BP <- -1 # Pa, negatve means altitude is used

# run subroutine
CONV.out <- CONV(TS, TENV, SHAPE, SURFAR, FLTYPE, FURTST, D, TFA, VEL, ZFUR, BP, ELEV)

QCONV <- CONV.out[1] # convective heat loss (W)
HC <- CONV.out[2] # combined convection coefficient
HCFREE <- CONV.out[3] # free convection coefficient
HCFOR <- CONV.out[4] # forced convection coefficient
HD <- CONV.out[5] # mass transfer coefficient
HDFREE <- CONV.out[6] # free mass transfer coefficient
HDFORC <- CONV.out[7] # forced mass transfer coefficient
ANU <- CONV.out[8] # Nusselt number (-)
RE <- CONV.out[9] # Reynold's number (-)
GR <- CONV.out[10] # Grasshof number (-)
PR <- CONV.out[11] # Prandlt number (-)
RA <- CONV.out[12] # Rayleigh number (-)
SC <- CONV.out[13] # Schmidt number (-)
BP <- CONV.out[14] # barometric pressure (Pa)

CONV.lab <- c("QCONV", "HC", "HCFREE", "HCFOR", "HD", "HDFREE", "HDFORC", "ANU", "RE", "GR", "PR", "RA", "SC", "BP")
kable(cbind(CONV.lab, t(CONV.out)))
```

### SEVAP

Compute evaporative heat exchange.

```{r}
BP <- BP # barometric pressure, Pa (from CONV)
TA <- 20 # air temperature (°C)
RELHUM <- 20 # relative humidity (%)
TC <- 37 # core temperature (°C)
TSKIN <- 33 # skin temperature (°C)
SKINW <- 11 # part of the skin surface that is wet (%)
FLYHR <- 0 # is flight occuring this hour? (imposes forced evaporative loss)
BAREVAP <- 0 # is evaporation partly from bare skin? (0 = no, 1 = yes, % defined with PCTSKINEVAP)
PCTBAREVAP <- 0 # surface area for evaporation that is skin, e.g. licking paws (%)
PCTEYES <- 0.03 # surface area made up by the eye (%) - make zero if sleeping
ZFUR <- ZFUR # fur depth, m
FURWET <- 0 # part of the fur surface that is wet (%)

SEVAP.out <- SEVAP(BP, TA, RELHUM, VEL, TC, TSKIN, ELEV, SKINW, FLYHR,  
  CONVSK, HD, HDFREE, PCTBAREVAP, PCTEYES, ZFUR, FURWET, TFA, CONVAR)

QSEVAP <- SEVAP.out[1] # skin evaporative heat loss (W)
WEYES <- SEVAP.out[2] # ocular evaporation (kg/s)
WCUTHF <- SEVAP.out[3] # forced cutaneous evaporation (kg/s)
WCUTF <- SEVAP.out[4] # free cutaneous evaporation (kg/s)
WCUT <- SEVAP.out[5] # total cutaneous evaporation (kg/s)
WTFUR <- SEVAP.out[5] # total fur evaporation (kg/s)
QFSEVAP <- SEVAP.out[5] # fur evaporative heat loss (W))

SEVAP.lab <- c("QSEVAP", "WEYES", "WCUTHF", "WCUTF", "WCUT", "WTFUR", "QFSEVAP")
kable(cbind(SEVAP.lab, t(SEVAP.out)))
```

### SIMULSOL

Setup and the SIMULSOL algorithm to SIMULtaneously SOLve for skin temperature $T_skin$ and the temperature of the fur/feather interface $T_fa$ given an environment, shape/posture and skin wetness. SIMULSOL is the core algorithm for computing the heat balance.

```{r}
# environment
FLTYPE <- 0 # FLUID TYPE: 0 = AIR; 1 = FRESH WATER; 2 = SALT WATER - need's to be looked at - only invoked in main program when the dive table is set upTC <- 37 # core temperature (°C)
TAREF <- TA # 1.2 m reference air temperature (°C)
TGRD <- TA # ground temperature (°C)
TSKY <- TA # sky temperature (°C)
TCONDSB <- TA # surface temperature for conduction (°C)
TBUSH <- TA # bush temperature (°C)
TVEG <- TAREF # assume vegetation casting shade is at 1.2 m (reference) air temperature (°C)
SKYIR <- C3 * (TSKY + 273.15) ^ 4 # sky infrared incoming (W)
VEGIR <- C6 * (TVEG + 273.15) ^ 4 # vegetation infrared incomming (W)
SKYRAD <- SKYIR + VEGIR 
SKYIN <- SKYRAD 
GRDIN <- C4 * (TGRD + 273.15) ^ 4 # note, MK put C4 here wherease before it was just SIG
TLOWER <- TGRD
RH <- 20 # relative humidity (%)
VEL <- 1 # wind speed (m s-1)
BP <- BP # Pa, negative means altitude is used (from CONV)
ELEV <- 0 # m

# physiology and morphology 
SKINW <- 0 # part of the skin surface that is wet (%)
AK1 <- 0.9 # initial thermal conductivity of flesh (0.412 - 2.8 W/mK)
AK2 <- 0.230 # conductivity of fat (W/mK)
BAREVAP <- 0 # is evaporation partly from bare skin? (0 = no, 1 = yes, % defined with PCTSKINEVAP)
PCTBAREVAP <- 0 # surface area for evaporation that is skin, e.g. licking paws (%)
PCTEYES <- 0.03 # surface area made up by the eye (%) - make zero if sleeping
# behaviour
FLYHR <- 0 # is flight occuring this hour? (imposes forced evaporative loss)

# configuration factors
FATOBJ <- 0 # configuration factor to nearby object
FAGRD <- 0.5 # configuration factor to ground
FASKY <- 0.5 # configuration factor to sky
FAVEG <- 0 # this is for overhead veg (at TAREF)
FABUSH <- 0 # this is for veg below/around animal (at TALOC)

# reference configuration factors
FABUSHREF <- FABUSH # nearby bush
FATOBJREF <- FATOBJ # nearby object
FASKYREF <- FASKY # sky
FAGRDREF <- FAGRD # ground
FAVEGREF <- FAVEG # vegetation

FURTHRMK <- 0 # user-specified fur thermal conductivity (W/mK), not used if 0
SHAPE <- 4 # shape, 1 is cylinder, 2 is sphere, 3 is plate, 4 is ellipsoid

# Initial values
TS <- TC # CURRENT GUESS OF OBJECT SURFACE TEMPERATURE
TFA <- TA # current guess of fur/air interface temperature

DIFTOL <- 0.001 # tolerance for SIMULSOL
SIMULSOL.out <- matrix(data = 0, nrow = 2, ncol = 15) # vector to hold the SIMULSOL results for dorsal and ventral side

# repeat for each side, dorsal and ventral, of the animal
for(S in 1:2){ 

# Calculating solar intensity entering fur. This will depend on whether we are calculating the fur temperature for the dorsal side or the ventral side. The dorsal side will have solar inputs from the direct beam hitting the silhouette area as well as diffuse solar scattered from the sky and objects. The ventral side will have diffuse solar scattered off the substrate.

# Setting config factors and solar depending on whether the dorsal side (S=1) or ventral side (S=2) is being estimated.
if(QSOLAR > 0.0){
  if(S==1){
    FASKY <- FASKYREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FATOBJ <- FATOBJREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FAVEG <- FAVEGREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FAGRD <- 0.0
    FABUSH <- 0.0
    if(FATOBJ == 0.0){
      QSLR <- 2*QSDIR+((QSSKY/FASKYREF)*FASKY)
    }else{
      QSLR <- 2*QSDIR+((QSSKY/FASKYREF)*FASKY)+((QSOBJ/FATOBJREF)*FATOBJ)
    }    
  }else{
    FASKY <- 0.0
    FATOBJ <- 0.0
    FAVEG <- 0.0
    FAGRD <- FAGRDREF/(1 - FAGRDREF - FATOBJREF - FAVEGREF)
    FABUSH <- FABUSHREF/(1 - FAGRDREF - FATOBJREF - FAVEGREF)
    QSLR <- QVENTR/(1 - FASKYREF - FATOBJREF - FAVEGREF)
  }
}else{
  QSLR <- 0.0 
  if(S==1){
    FASKY <- FASKYREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FATOBJ <- FATOBJREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FAVEG <- FAVEGREF/(FASKYREF+FATOBJREF+FAVEGREF)
    FAGRD <- 0.0
    FABUSH <- 0.0
  }else{
    FASKY <- 0.0
    FATOBJ <- 0.0
    FAVEG <- 0.0
    FAGRD <- FAGRDREF/(1 - FAGRDREF - FATOBJREF - FAVEGREF)
    FABUSH <- FABUSHREF/(1 - FAGRDREF - FATOBJREF - FAVEGREF)
  }
}
# set fur depth and conductivity
# index for KEFARA, the conductivity, is the average (1), front/dorsal (2), back/ventral(3) of the body part
if(QSOLR > 0 | ZFURD != ZFURV){
  if(S == 1){
    ZL <- ZFURD
    KEFF <- KEFARA[2] 
  }else{
    ZL <- ZFURV
    KEFF <- KEFARA[3]
  }
}else{
  ZL <- ZFUR
  KEFF <- KEFARA[1]
}

RDXDEP <- 1 # not used yet - relates to radiation through fur
XR <- RDXDEP # not used yet - relates to radiation through fur
X <- RDXDEP # not used yet - relates to radiation through fur
RSKIN <- R1 # body radius (including fat), m
RFLESH <- R1 - FATTHK # body radius flesh only (no fat), m
RFUR <- R1 + ZL # body radius including fur, m
D <- 2 * RFUR # diameter, m
RRAD <- RSKIN + (XR * ZL) # effective radiation radius, m
LEN <- ALENTH # length, m

# Correcting volume to account for subcutaneous fat
if(SUBQFAT == 1 & FATTHK > 0.0){
  VOL <- FLSHVL
}

# Getting compressed fur thermal conductivity
AREACND <- ATOT * PCOND
CD <- AREACND * ((KFURCMPRS / ZFURCOMP))

# package up inputs
FURVARS <- c(LEN, ZFUR, FURTHRMK, KEFF, BETARA, FURTST, ZL)
GEOMVARS <- c(SHAPE, SUBQFAT, CONVAR, VOL, D, CONVAR, CONVSK, RFUR, RFLESH, RSKIN, XR, RRAD, ASEMAJ, BSEMIN, CSEMIN, CD)
ENVVARS <- c(FLTYPE, TA, TS, TBUSH, TVEG, TLOWER, TSKY, TCONDSB, RH, VEL, BP, ELEV, FASKY, FABUSH, FAVEG, FAGRD, QSLR)
TRAITS <- c(TC, AK1, AK2, EMISAN, FATTHK, FLYHR, FURWET, PCTBAREVAP, PCTEYES)

# set IPT, the geometry assumed in SIMULSOL: 1 = cylinder, 2 = sphere, 3 = ellipsoid
if(SHAPE %in% c(1,3,5)){
  IPT <- 1
}
if(SHAPE == 2){
  IPT <- 2
}
if(SHAPE == 4){
  IPT <- 3
}

# call SIMULSOL
SIMULSOL.out[S,] <- SIMULSOL(DIFTOL, IPT, FURVARS, GEOMVARS, ENVVARS, TRAITS, TFA, SKINW, TS)
}

SIMULSOL.out <- cbind(c(1,2), SIMULSOL.out)
colnames(SIMULSOL.out) <- c("SIDE", "TFA", "TSKIN", "QCONV", "QCOND", "QGENNET", "QSEVAP", "QRAD", "QSLR", "QRSKY", "QRBSH", "QRVEG", "QRGRD", "QFSEVAP", "NTRY", "SUCCESS")
tSIMULSOL.out <- t(SIMULSOL.out)
colnames(tSIMULSOL.out) <- c("DORSAL", "VENTRAL")
kable(tSIMULSOL.out)
```

### ZBRENT and RESPFUN

Finally, find a value of QRESP (heat lost by respiration) to balance the heat budget by solving RESPFUN with ZBRENT.

```{r}
# define basal metabolic rate
QBASAL <- (70 * AMASS ^ 0.75) * (4.185 / (24 * 3.6)) # heat generation (W)
DELTAR <- 0 # offset between air temeprature and breath (°C)
O2GAS <- 20.95 # oxygen concentration of air (%)
N2GAS <- 79.02 # nitrogen concetration of air (%)
CO2GAS <- 0.03 # carbon dioxide concentration of air (%)
RQ <- 0.80 # respiratory quotient (fractional, typically 0.7 (fats) to 1 (carbs), with 0.8 typical for protein)
EXTREF <- 20 # O2 extraction efficiency (%)
RELXIT <- 100 # relative humidity of exhaled air (%)
TIMACT <- 1 # multiplier on metabolic rate for activity costs
PANT <- 1 # multiplier on breathing rate (-)

# Now compute a weighted mean heat generation for all the parts/components = (dorsal value *(FASKY+FAVEG+FATOBJ))+(ventral value*FAGRD)
GEND <- SIMULSOL.out[1, 6]
GENV <- SIMULSOL.out[2, 6]
DMULT <- FASKYREF + FAVEGREF + FATOBJ
VMULT <- 1 - DMULT # Assume that reflectivity of veg below = ref of soil so VMULT left as 1 - DMULT
X <- GEND * DMULT + GENV * VMULT # weighted estimate of metabolic heat generation

# lung temperature and temperature of exhaled air
TLUNG <- (TC + (SIMULSOL.out[1, 3] + SIMULSOL.out[1, 3]) * 0.5) * 0.5 # average of skin and core
TAEXIT <- min(TA + DELTAR, TLUNG) # temperature of exhaled air, °C

QMIN <- QBASAL
QM1 <- X - (5 * QMIN) 
QM2 <- X + (10 * QMIN)
QSUM <- X
TOL <- AMASS * 0.01

ZBRENT.in <- c(TA, O2GAS, N2GAS, CO2GAS, BP, QMIN, RQ, TLUNG, GMASS, EXTREF, RH, RELXIT, TIMACT, TAEXIT, QSUM, PANT)

# call ZBRENT subroutine which calls RESPFUN
ZBRENT.out <- ZBRENT(QM1, QM2, TOL, ZBRENT.in)

colnames(ZBRENT.out) <- c("RESPFN","QRESP","GEVAP", "PCTO2", "PCTN2", "PCTCO2", "RESPGEN", "O2STP", "O2MOL1", "N2MOL1", "AIRML1", "O2MOL2", "N2MOL2", "AIRML2", "AIRVOL")
tZBRENT.out <- t(ZBRENT.out)
colnames(tZBRENT.out) <- c("OUTPUT")
kable(tZBRENT.out)
```
