---
title: "endoR function examples"
author: "Michael Kearney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{endoR function examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 eval = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE  
)
```

## Overview

This document shows how to use the function `solve.endo.R`, the main part of the 'endoR' collection of routines in NicheMapR, to compute endotherm heat and mass balances. This function collates all the different steps in [endoR example](endoR-example.html) into a single function that can be applied to vectors of data. The `solve.endo.R` function has one behavioural option at present - sweating to cool.

## Single calculation example

Here is how you run the calculation with all default values except specifying a local air temperature of 0 deg C.

```{r}
endo.out <- solve.endo(TA = 0)
kable(endo.out)
```

## Metabolic chamber example

Now try a sequence of air temperatures for a specified core temperature, mass, fur depth and base skin wetness, with the option to sweat at 0.1 % increments to cool down if needed. Note that the default settings make the environment a blackbody one (all surface and air temperatures the same) with low (5%) relative humidity and low wind speed (0.1 m/s), as is often the case in metabolic chamber experiments.

```{r, fig.width = 7, fig.height = 5}
TAs <- seq(0,50) # air temperature (deg C)
TC <- 37 # core temperature (deg C)
AMASS <- 1 # mass (kg)
ZFURD <- 2E-03 # fur depth, dorsal (m)
ZFURV <- 2E-03 # fur depth, dorsal (m)
GMREF <- 4
SKINW <- 0.05
SWEAT <- .1
QBASAL <- (70 * AMASS ^ 0.75) * (4.185 / (24 * 3.6)) # basal heat generation (W)

ptm <- proc.time()
endo.out <- lapply(1:length(TAs), function(x){solve.endo(TA = TAs[x], AMASS = AMASS, ZFURD = ZFURD, ZFURV = ZFURV, GMREF = GMREF, SKINW = SKINW, SWEAT = SWEAT, QBASAL = QBASAL)})
proc.time() - ptm
endo.out <- do.call("rbind", lapply(endo.out, data.frame))

QGEN <- endo.out$RESPGEN # metabolic rate (W)
H2O <- endo.out$EVAP.G.H # g/h water evaporated
TFA_D <- endo.out$TFA_D # dorsal fur surface temperature
TFA_V <- endo.out$TFA_V # ventral fur surface temperature
TskinD <- endo.out$TSKIN_D # dorsal skin temperature
TskinV <- endo.out$TSKIN_V # ventral skin temperature
TC <- endo.out$TC # core temperature
plot(QGEN ~ TAs, type = 'l', ylab = 'metabolic rate, W', xlab = 'air temperature, deg C')
plot(H2O ~ TAs, type = 'l', ylab = 'water loss, g/h', xlab = 'air temperature, deg C')
plot(TFA_D ~ TAs, type = 'l', col = 'grey', ylab = 'fur, skin and core temperature, deg C', xlab = 'air temperature, deg C')
points(TFA_V ~ TAs, type = 'l', col = 'grey', lty = 2)
points(TskinD ~ TAs, type = 'l', col = 'orange')
points(TskinV ~ TAs, type = 'l', col = 'orange', lty = 2)
points(TC ~ TAs, type = 'l', col = 'red')
```

## Microclimate model integration example

In this final example, the microclimate model is run and the then the relevant outputs are sent into the endotherm model along with the same animal parameters as above.

```{r, fig.width = 7, fig.height = 5}
micro <- micro_global(loc = c(145, -37), runshade = 0, Usrhyt = 0.01)

metout <- as.data.frame(micro$metout)
soil <- as.data.frame(micro$soil)
days<-rep(seq(1,12),24)
days<-days[order(days)]
dates<-days+metout$TIME/60/24-1 # dates for hourly output

TAs <- metout$TALOC
TAREFs <- metout$TAREF
TSKYs <- metout$TSKYC
TGRDs <- soil$D0cm
VELs <- metout$VLOC
RHs <- metout$RHLOC
QSOLRs <- metout$SOLR
Zs <- metout$ZEN
ELEV <- micro$elev
ABSSB <- 1-micro$REFL


endo.out <- lapply(1:length(TAs), function(x){solve.endo(TA = TAs[x], TAREF = TAREFs[x], TSKY = TSKYs[x], 
  TGRD = TGRDs[x], VEL = VELs[x], RH = RHs[x], QSOLR = QSOLRs[x], Z = Zs[x], ELEV = ELEV, ABSSB = ABSSB, AMASS = AMASS, ZFURD = ZFURD, ZFURV = ZFURV, GMREF = GMREF, SKINW = SKINW, SWEAT = SWEAT)})
endo.out <- do.call("rbind", lapply(endo.out, data.frame))

QGEN <- endo.out$RESPGEN # metabolic rate (W)
H2O <- endo.out$GEVAP * 3600 # g/h water evaporated
TFA_D <- endo.out$TFA_D # dorsal fur surface temperature
TFA_V <- endo.out$TFA_V # ventral fur surface temperature
TskinD <- endo.out$TSKIN_D # dorsal skin temperature
TskinV <- endo.out$TSKIN_V # ventral skin temperature
TC <- endo.out$TC # core temperature

plot(QGEN ~ dates, type = 'l', ylab = 'metabolic rate, W', xlab = 'time')
plot(H2O ~ dates, type = 'l', ylab = 'water loss, g/h', xlab = 'time')
plot(TFA_D ~ dates, type = 'l', col = 'grey', ylab = 'fur, skin and core temperature, deg C', xlab = 'time')
points(TFA_V ~ dates, type = 'l', col = 'grey', lty = 2)
points(TskinD ~ dates, type = 'l', col = 'orange')
points(TskinV ~ dates, type = 'l', col = 'orange', lty = 2)
points(TC ~ dates, type = 'l', col = 'red')
```
