---
title: "Underlying theory and equations of the Niche Mapper microclimate model"
author: "Michael Kearney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the microclimate model of NicheMapR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Model Structure

The microclimate model implemented in NicheMapR (originally described in Beckman et al., 1973, Mitchell et al., 1975, Porter et al., 1973, McCullough & Porter, 1971) includes integrated routines for calculating solar and infra-red radiation intensities, above-ground profiles of air temperature, wind velocity, and relative humidity, and soil temperature profiles from the surface down to a user-specified maximum depth (typically around 2m). The model has more recently been developed to include a full water balance and snow. These calculations are made for two user-provided extremes of shading by vegetation.

## Radiation Model

Solar radiation has a dominating effect on the available microclimates across the earth. The total solar irradiance (W/m2) reaching the earth at a given location depends on 
1.	the radiation reaching the outer edge of the atmosphere at the location of interest (extraterrestrial radiation – affected by the earth’s orbit, time of year and day, and location on the earth);
2.	the effect of the atmosphere as the radiation travels through it to the ground (terrestrial radiation – affected by attenuation due to Raleigh scattering, aerosols, water vapour, ozone and cloud);
3.	the terrain (vegetation shade, hillshade, slope, aspect). 
The subroutine used in NicheMapR for calculating clear sky solar radiation is based on McCullough and Porter (1971) with modifications to incorporate skylight before sunrise and after sunset (Rozenberg, 1966, p. 19) and to include the effects of sloping surfaces and hillshade. 

## Extraterrestrial Radiation

The total solar irradiance reaching an imaginary plane perpendicular to the sun’s rays at the outer edge of the atmosphere is obtained by integrating across all wavelengths of radiation emitted by the sun (290 to 4000 nm) after correcting for the distance of the earth from the sun and for the angle at which the sun is shining on that plane. Thus, the wavelength-specific irradiance reaching this plane



Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
