---
title: "Properties of Air: A Manual for Use in Biophysical Ecology, 4th Edition"
author: "C. Richard Tracy, William R. Welch, Michael R. Kearney and Warren P. Porter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Properties of Air}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 eval = TRUE
)
```
This manual was made possible by the invaluable assistance of Ann Chambers and Cheryle Hughes. Financial support was provided by grants from the Department of Zoology, Wisconsin Alumni Research Foundation, ERDA (Contract EY-76-5-02-2270), and NSF (Grant Nos. 74-19454 and 77-25786). Michael Kearney updated the document, translated the Fortrans subroutines DRYAIR and WETAIR, and the function VAPPRS into R, and made it a vignette as part of the *NicheMapR* package.  

<center>First edition 1973  
Second edition 1978  
Third edition 1980  
Fourth edition 2016  </center>

## Introduction

This manual comprises a series of tables and graphs that illustrate selected properties of air as functions of temperature, pressure, and humidity. The particular properties that are displayed were chosen because of their importance in analytical studies of energy (heat) and mass (water) transfer between organisms and their physical environments. Except where otherwise noted, the graphs were drawn from information in the Smithsonian Meteorological Tables (List 1971). These graphs are considered to be a concise references for the values of the indicated properties, and not substitutes for the more accurate Smithsonian tables.

All information in this manual is expressed in terms of the International System of Units (SI), which is based, in part, on the metre, kilogram, second and kelvin. Basic units, derived units, prefixes, and some useful physical constants and conversion factors are listed below. More extensive lists are provided by List (1971), Mechtly (1973). and the Symbols Committee of the Royal Society (1975). Note that the symbol for a physical quantity is an italicized (slanted) letter of the Roman or Greek alphabet, and the unit for a quantity is indicated by an unitalicized (upright) Roman letter.
An equation is provided as a mathematical description of each graph. In addition, two Fortran subroutines that calculate all of the information in the graphs are listed. Two of the equations in subroutine DRYAIR (thermal conductivity of air and latent heat of vaporization of water) are linear regressions that were fit to tabular data. All other equations in DRYAIR were taken from the Smithsonian tables.

All humidity calculations in subroutine WETAIR are made with the internationally accepted Goff-Gratch equation (function VAPPRS in subroutine WETAIR). A mathematically less cumbersome and often more useful alternative was reported by Tetens (Murray 1976); calculations with this equation differ negligibly from those with the Goff-Gratch expression. Teten’s equation is provided on page x to calculate the saturation vapor pressure ($e_d^*$) at the dewpoint temperature ($t_d$). Note that this equation can also be used to calculate the saturation vapor pressure ($e^*$) at the dry bulb temperature ($t$) if t replaces td, and the saturation vapor pressure ($e_w^*$) at the wet bulb temperature ($t_w$) if $t_w$ replaces $t_d$. The relationship of wet bulb temperature to dry bulb temperature, vapor density, and relative humidty is shown in the psychometric diagram on page x.

Mathematical expressions for some properties of air that are dependent on humidity are not available. The equations for these properties are restricted to dry air in this manual; the accompanying graphs, however, include comparisons for saturated air, shown by broken lines, from the theoretical calculations of Mason and Monchick (1965).


### Table 1. Basic units
```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Physical Quantity*       |	*Quantity Symbol* |	*Unit*   |	*Unit Symbol*
------------------------- | ----------------- | -------- | --------------
length                    | *l*               | metre    | m
mass                      | *m*               | kilogram | kg
time                      | *t*               | second   | s
thermodynamic temperature | *T*               | kelvin   | K
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

### Table 2. Derived units
```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Physical Quantity*  | *Quantity Symbol*   | *Unit*   | *Unit Symbol*   | *Unit Definition*
-------------------- | ------------------- | -------- | --------------- | -----------------
force                | *F*                 | newton   | N               | $J$ $m^{-1}$
energy               | *E*                 | joule    | J               | $N$ $m$
pressure             | *p*                 | pascal   | Pa              | $N$ $m^{-2}$, $J$ $m^{-3}$
power                | *P*                 | watt     | W               | $J$ $s^{-1}$
volume               | *t*                 | cubic metre | $m^3$        | $m^3$
Celsius temperature$^*$ | $t$              | degree Celsius   | &deg;C   | $K$
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

### Table 3. Prefixes
```{r table3, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Multiple* | *Prefix*	| *Symbol* |	*Multiple* | *Prefix*	 | *Symbol*
---------- | -------- | -------- | ----------- | --------- | -------
$10^{-1}$	 | deci	    | *d*      | $10^{1}$	   | deca	     | da
$10^{-2}$	 | centi    | *c*      | $10^{2}$	   | hecto     | h
$10^{-3}$	 | milli    | *m*      | $10^{3}$	   | kilo	     | k
$10^{-6}$	 | micro    | *μ*      | $10^{6}$	   | mega	     | M
$10^{-9}$	 | nano	    | *n*      | $10^{9}$	   | giga	     | G
$10^{-12}$ | pico	    | *p*      | $10^{12}$   | tera	     | T
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```


### Table 4. Conversion factors
```{r table4, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
**Force**   
&nbsp;&nbsp;&nbsp;&nbsp;dyne = $10^5$ N$^*$  
&nbsp;&nbsp;&nbsp;&nbsp;kilogram force = $9.80665$ N$^*$  
&nbsp;&nbsp;&nbsp;&nbsp;ounce force = $2.7801385$ x $10^{-1}$ N  
&nbsp;&nbsp;&nbsp;&nbsp;pound force = $4.4482216152605$ N$^*$  
   
**Energy**   
&nbsp;&nbsp;&nbsp;&nbsp;British thermal unit (thermochemical) = $1.054350$ x $10^{3}$ J  
&nbsp;&nbsp;&nbsp;&nbsp;calorie (thermochemical) = $4.184$ J  
&nbsp;&nbsp;&nbsp;&nbsp;what hour = $3.60$ x $10^{3}$ J$^*$  
   
**Pressure**   
&nbsp;&nbsp;&nbsp;&nbsp;bar = $10^{5}$ $Pa^*$  
&nbsp;&nbsp;&nbsp;&nbsp;centimetre of water (4 &deg;C) = $9.80638$ x $10^{1}$ Pa   
&nbsp;&nbsp;&nbsp;&nbsp;millimetre of mercury (0 &deg;C) = $1.333224$ x $10^{2}$ Pa   
&nbsp;&nbsp;&nbsp;&nbsp;pound per square inch = $6.894757$ x $10^{3}$ Pa   
&nbsp;&nbsp;&nbsp;&nbsp;standard atmosphere = $1.01325$ x $10^{5}$ Pa   
&nbsp;&nbsp;&nbsp;&nbsp;torr = $1.333224$ x $10^{2}$ Pa   
   
**Power**   
&nbsp;&nbsp;&nbsp;&nbsp;British thermal unit (thermochemical) = $1.054350$ x $10^{3}$ W   
&nbsp;&nbsp;&nbsp;&nbsp;calorie (thermochemical) per second = $4.184$ W$^*$   
&nbsp;&nbsp;&nbsp;&nbsp;calorie (thermochemical) per minute = $6.973333$ x $10^{-2}$ W   
&nbsp;&nbsp;&nbsp;&nbsp;kilocalorie (thermochemical) per hour = $8.604207$ x $10^{-1}$ W   
   
**Volume**   
&nbsp;&nbsp;&nbsp;&nbsp;cubic foot = $2.8316592$ m$^{3*}$   
&nbsp;&nbsp;&nbsp;&nbsp;gallon = $3.785411784$ x $10^{-3}$ m$^{-3*}$    
&nbsp;&nbsp;&nbsp;&nbsp;litre = $10^{-3}$ m$^{-3*}$   
&nbsp;&nbsp;&nbsp;&nbsp;pint (U.S. liquid) = $4.73176473$ x $10^{-4}$  m$^{3*}$   
&nbsp;&nbsp;&nbsp;&nbsp;quart (U.S. liquid) = $9.4635295$ x $10^{-4}$  m$^{3*}$  
   
**Length**   
&nbsp;&nbsp;&nbsp;&nbsp;foot = $3.048$ x $10^{-1}$ m$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;inch = $2.54$ x $10^{-2}$ m$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;mile (U.S. statute) = $1.609344$ x $10^{3}$ m$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;mile (U.S. nautical) = $1.852$ x $10^{3}$ m$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;yard = $9.144$ x $10^{-1}$ m$^{*}$   
   
**Mass**
&nbsp;&nbsp;&nbsp;&nbsp;ounce = $2.8349523125$ x $10^{-2}$ kg$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;pound = $4.5359237$ x $10^{-1}$ kg$^{*}$   
&nbsp;&nbsp;&nbsp;&nbsp;ton (short, 2000 pound) = $9.0718474$ x $10^{2}$ kg$^{*}$   
   
$^{*}$exact, by definition
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

### Table 5. Physical constants
```{r table5, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Quantity*                                           | *Symbol*	| *Value*  |	*Units*
---------------------------------------------------- | -------- | -------- | -----------
Density of dry air at 25 &deg;C and 101325 Pa$^{*}$	 | &rho;	    | $1.184$    | kg m$^3$	   
Diffusivity of water vapour in air at 25 &deg;C and 101325 Pa$^{*}$	 | *D*	    | $2.614$ x $10^{-5}$    | m$^2$ s$^{-1}$	   
Dynamic viscosity of dry air at 25 &deg;C and 101325 Pa$^{*}$	 | &mu;	    | $1.806$ x $10^{-5}$    | kg m$^{-1}$ s$^{-1}$	   
Gas constant for ideal gas	 | *R*	    | $8.31434$    | J mol$^{-1}$ K$^{-1}$   
Gas constant for water vapour	 | *R*$_v$	    | $4.6150$ x $10^{2}$    | Pa m$^3$ kg$^{-1}$ K$^{-1}$	   
Gravitational constant	 | *R*	    | $6.6732$ x $10^{-11}$    | N m$^2$ kg$^{-2}$	   
Kinematic viscosity of dry air at 25 &deg;C and 101325 Pa$^{*}$	 | &nu;	    | $1.525$ x $10^{-5}$    | m$^3$ s$^{-1}$	
Latent heat of vapourisation of water at 25 &deg;C	 | *L*	    | $2.442$ x $10^{6}$    | J kg$^{-1}$	   
Molecular weight of dry air	 | *M*$_a$	    | $2.8966$ x $10^{-2}$    | kg mol$^{-1}$	   
Molecular weight of water | *M*$_v$	    | $1.8016$ x $10^{-2}$    | kg mol$^{-1}$	 
Specific heat of dry air	 | *c*$_p$	    | $1.00484$ x $10^{3}$    | K kg$^{-1}$ K$^{-1}$	   
Standard pressure	 | *P*$_0$ 	    | $1.01325$ x $10^{5}$    | Pa	   
Standard temperature	 | *T*$_0$ 	    | $2.7315$ x $10^{2}$    | K	   
Stefan-Boltzmann constant	 | &sigma;	    | $5.67032$ x $10^{-8}$    | W m$^{-2}$ K$^{-4}$	   
Thermal conductivity of dry air at 25 &deg;C and 101325 Pa$^{*}$	 | *k*	    | $2.601$ x $10^{-2}$    | W m$^{-1}$ K$^{-1}$	   
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{*}$Variable that is functionally related to other properties in this manual

### Table 6. Percent relative humidity over saturated solutions as a function of temperature. From Winston and Bates (1960)$^1$
```{r table6, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "

*Compound* | *10&deg;C*	| *15&deg;C*  |	 *20&deg;C* | *25&deg;C*	| *30&deg;C*	| *35&deg;C*	
--- | --- | --- | --- | --- | --- | --- 
NH$_4$Cl|79.0 | 79.5 | 79.5 | 78.0 | 77.5 | -
NH$_4$HPO$_4$ (monophosphate)|- | - | 93.0 | 93.0 | 92.0 | -
NH$_4$NO$_3$|75.0 | 70.0 | 65.5 | 62.5 | 59.5 | 55.0
NH$_4$NO$_3$ + NaNO$_3$|58.0 | 55.0 | 52.0 | 50.0 | 47.0 | 44.5
NH$_4$NO$_3$ + AgNO$_3$|70.5 | 68.0 | 65.1 | 61.5 | 58.0 | 55.0
(NH$_4$)$_2$ SO$_4$|80.5 | 81.0 | 80.5 | 80.0 | 80.0 | 79.5
CaCl$_2$&middot;6H$_2$O$^2$|38.0 | 35.0 | 32.5 | 29.5 | - | -
CaHPO$_4$&middot;2H$_2$O$^3$|97.5 | 99.5 | 95.0 | 97.0 | 95.0 | -
CaH$_4$(PO$_4$)&middot;H$_2$O$^4$|98.0 | 99.0 | 94.0 | 96.0 | 93.5 | -
Ca(NO$_3$)$_2$&middot;4H$_2$O$^5$|- | 56.0 | 55.5 | 50.5 | 47.0 | -
C0Cl$_2$$^6$|- | 72.5 (18 &deg;C) | 67.0 | - | 62.0 | -
Cr$_2$O$_3$$^7$|- | 45.5 | (39.0) | - | 44.5 | -
Glucose|57 (12 &deg;C) | - | 55.0 | 55.0 | - | 53.0
LiCl&middot;H$_2$O$^8$|13.5 | 13.0 | 12.5 | 12.0 | 11.5 | 11.5
Pb(NO$_3$)$_2$$^9$|98.0 | 97.0 | 97.0 | 95.5 | 95.0 | 94.5
Pb(NO$_3$)$_2$+NH$_4$NO$_3$|66.5 | 62.0 | 58.0 | 55.0 | 52.5 | 49.5
MgCl$_2$&middot;6H$_2$O|34.0 | 34.0 | 33.0 | 32.5 | 32.5 | 32.5
Mg(NO$_3$)$_2$&middot;6H$_2$O|58.0 | 56.0 | 55.0 | 53.0 | 52.0 | 50.5
P$_2$O$_5$|0.0 | 0.0 | 0.0 | 0.0 | 0.0 | -
KAc|(21.0) | - | 20.0 | 22.5 | 22.0 | -
KBr|86.0 | - | 84.0 | 80.0 | 82.0 | -
K$_2$CO$_3$&middot;2H$_2$O$^{10}$|47.0 | 44.0 | 44.0 | 43.0 | 43.0 | -
KCl|88.0 | 86.5 | 85.0 | 85.0 | 84.5 | 83.0
K$_2$CrO$_4$|- | - | 88.0 | - | 86.5 | -
K$_2$Cr$_2$O$_7$|- | - | 98.0 | 98.0 | 97.5 | 96.5
KH$_2$PO$_4$|98.0 | 99.0 | 96.5 | 96.0 | 93.5 | -
KF&middot;2H$_2$O$^{11}$|- | - | - | 30.5 | 27.4 | -
KNO$_3$$^{12}$|96.0 | 95.5 | 93.5 | 92.5 | 91.0 | 89.5
KNO$_2$|- | - | 48.5 | - | 47.0 | -
K$_2$SO$_4$|98.5 | 99.0 | 98.0 | 97.5 | 96.5 | 96.0
KCNS$^{13}$|52.0 | 50.0 | 47.0 | 46.5 | 43.5 | 41.5
K tartrate$^{14}$|- | 75.0 | 75.0 | 75.0 | 74.0 | -
KNa tartrate$^{15}$|87.5 | - | 87.0 | 87.0 | 87.0 | -
Pyrocatechol|- | 99.0 | 95.5 | 93.5 | 92.5 | 90.5
Resorinol|- | 95.0 | 87.0 | 85.0 | 82.5 | 79.5
AgNO$_3$|88.0 | 86.0 | 84.0 | 82.0 | 80.0 | 78.0
AgNO$_3$+Pb(NO$_3$)$_2$|84.0 | 83.0 | 80.5 | 78.5 | 76.5 | 75.0
NaAc&middot;3H$_2$O|- | - | 75.0 | 73.0 | 71.0 | -
NaBr&middot;2H$_2$O$^{16}$|63.0 | 61.0 | 59.0 | 57.5 | 56.0 | 54.5
Na$_2$CO$_3$&middot;10H$_2$O$^{17}$|(99.0) | - | 92.0 | 87.0 | 87.0 | -
NaCl|76.5 | 76.0 | 76.0 | 75.5 | 75.5 | 75.5
NaCl+KCl|- | - | 70.0 | 71.5 | 71.0 | -
Na$_2$CrO$_4$&middot;4H$_2$O|- | - | - | - | 64.5 | -
Na$_2$Cr$_2$O$_7$&middot;H$_2$O$^{18}$|60.0 | 56.5 | 54.5 | 53.0 | 52.5 | 51.0
NaOH|(5.5) | - | 5.5 | 7.0 | (4.0) | -
NaI|- | - | 38.0 | 38.0 | 36.0 | -
NaNO$_3$|77.5 | 76.5 | 76.0 | 74.0 | 72.5 | 71.0
NaNO$_2$|- | - | 65.5 | 64.0 | 63.0 | -
Na$_2$SO$_4$&middot;10H$_2$O|- | - | (93.0) | (93.0) | - | (87.0)
Na tartrate$^{19}$|- | 94.0 | 92.0 | 92.0 | 92.0 | -
Urea|81.5 | 80.0 | 80.0 | 76.0 | 73.3 | -
ZnCl$_2$&middot;1.5H$_2$O$^{20}$|(10.0) | (10.0) | 10.0 | - | (10.0) | -
ZnSO$_4$&middot;7H$_2$O$^{21}$|- | - | - | 88.5 | - | -
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Parentheses around a value of relative humidity indicate some doubt of accuracy.   
$^{2}$May be unstable around 20 &deg;C.   
$^{3}$May be a transition point between 15 and 20 &deg;C.   
$^{4}$Calcium monophosphate.   
$^{5}$Good up to 40 &deg;C.   
$^{6}$Uncertain around 20 &deg;C, large temperature coefficient.   
$^{7}$Other researchers report 40\% at 30 &deg;C.   
$^{8}$LiCl&middot;XH$_2$O &rarr; LiCl&middot;H$_2$O at 25 &deg;C.     
$^{9}$Solution slightly acid due to hydrolysis, but stable.   
$^{10}$Should be granulated then moistened; powdered from is unstable.   
$^{11}$KF&middot;2H$_2$O &rarr; KF at 50 &deg;C.   
$^{12}$Large temperature coefficient.   
$^{13}$Very toxic.   
$^{14}$Potassium tartrate. Excellent.   
$^{15}$Potassium sodium tartrate (Rochelle salts). Good only below 40 &deg;C.   
$^{16}$Good up to 55 &deg;C, large temperature coefficient.   
$^{17}$NaCO$_3$&middot;10H$_2$O.....>Na$_2$CO$_3$&middot'H$_2$O at 32 &deg;C.   
$^{18}$Very good; small temperature coefficient.   
$^{19}$This is also listed with one water of hydration, instead of two; good only below 40 &deg;C.   
$^{20}$Unstable around 20 &deg;C.   
$^{21}$Good below 10 &deg;C.      

### Table 7. Percent relative humidity over saturated solutions at 20 &deg;C. From Winston and Bates (1960)
```{r table7, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "

*Compound* | *Relative Humidity*	| *Compound*  |	 *Relative Humidity*	
--- | --- | --- | --- 
BaI$_2$&middot;2H$_2$O|44.0 | NaHSO$_4$ | 52.0
CaSO$_4$&middot;2H$_2$O|98.0 | NaBrO$_3$ | 92.0
NH$_4$&middot;4H$_2$O|65.0 | NaNO$_3$+KNO$_3$| 71.0
NH$_4$&middot;7H$_2$O|90.0 | Na$_2$HPO$_4$&middot;7H$_2$O | 95.0
NH$_4$&middot;4H$_2$O|57.5 | Na$_2$SO$_3$$^*$ | 95.0
NH$_4$&middot;6H$_2$O|55.5 | NaCNS | 35.5
H$_3$PO$_4$|9.0 | Na$_2$S$_2$O$_3$ | 78.0
KHSO$_4$|86.0 | 2n(NO$_3$)$_2$ | 42.0
K$_2$HPO$_4$|44.5 | - | -
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{*}$Good only below 29.6 &deg;C.

### Table 8. Percent relative humidity over saturated solutions at 25 &deg;C. From Winston and Bates (1960)$^1$
```{r table8, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "

*Compound* | *Relative Humidity*	| *Compound*  |	 *Relative Humidity*	
--- | --- | --- | --- 
NH$_4$Br$_2$|75.0 | MgSO$_4$&middot;7H$_2$O | 89.0
BaBr$_2$|74.5 | Mg(CNS)$_2$ | 47.5
BaCl$_2$|90.0 | MnBr$_2$&middot;6H$_2$O | 34.5
BaI$_2$&middot;2H$_2$O|43.0 | MnCl$_2$&middot;4H$_2$O | 56.0
Ba(CNS)$_2$&middot;2H$_2$O|54.5 | NiBr$_2$&middot;3H$_2$O | 27.0
CaAc$_2$&middot;H$_2$O|17.0 | NiCl$_2$&middot;6H$_2$O | 53.0
CaAc$_2$&middot;H$_2$O + sucrose|13.0 | H$_3$PO$_4$ | 9.0
CaBr$_2$|16.5 | KBr + sucrose | 63.0
CaI$_2$&middot;6H$_2$O|11.5 | KBr + Urea | 51.0
Ca methane sulphonate|72.5 | KClO$_3$ | 98.0
Ca(MnO$_4$)$_2$&middot;4H$_2$O|37.5 | KCl + KClO$_3$ | 85.0
Ca(CNS)$_2$&middot;3H$_2$O|17.5 | KHCO$_4$ | 21.5
CaZnCl$_2$|20.0 | KOH | 8.0
CeCl$_3$ (tech.)|45.5 | K$_4$P$_2$O$_7$&middot;3H$_2$O | 49.5
CrCl$_3$|42.5 | K$_2$S$_2$O$_3$ | 66.0
CoBr$_2$|41.5 | NaAc + sucrose | 37.6
Co(NO$_3$)$_2$&middot;6H$_2$O|(49.0) | NaCl + sucrose | 63.0
CuBr$_2$|62.5 | NaCl + NaNO$_3$ | 69.0
CuCl$_2$&middot;2H$_2$O|67.0 | NaCl + Na$_2$SO$_4$&middot;7H$_2$O | 74.0
Cu(NO$_3$)$_2$&middot;6H$_2$O|(35.0) | NaHSO$_4$ | 81.0
Ethanolamine sulphate|12.5 | NaH$_2$PO$_4$ | 70.0
FeBr$_2$&middot;6H$_2$O|39.0 | Na methane sulphonate | 61.5
FeCl$_2$&middot;4H$_2$O|60.0 | SrBr$_2$&middot;6H$_2$O | 58.5
LiBr&middot;2H$_2$O|7.0 | SrI$_2$&middot;6H$_2$O | 33.0
LiI&middot;3H$_2$O|18.0 | Sr(CNS)$_2$&middot;3H$_2$O | 31.5
LiNo$_3$&middot;3H$_2$O|47.0 | Sucrose | 85.0
LiCNS$^2$|12.5 | Surcose + urea | 52.0
MgBr$_2$&middot;6H$_2$O|31.5 | ZnBr$_2$ | 8.5
MgI$_2$|27.0 | ZbI$_2$$^2$ | 20.0
Mg(ClO$_4$)$_2$&middot;6H$_2$O|(41.0) | Zn(MnO$_4$)$_2$&middot;6H$_2$O | 51.0
Mg silicoflouride|91.5 | Zn(CNS)$_2$ | 80.5
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Parentheses around a value of relative humidity indicate some doubt of accuracy.   
$^{2}$Light sensitive.   
$^{3}$Prepare by mixing Ca(OH)$_2$ + ZnO + 4HCl.   

### Table 9. Efficiency of drying agents expressed as mass of water remaining per unit volume of air dried at 25 &deg;C, and as relative humidity$^*$. From Winston and Bates (1960).
```{r table9, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "

*Compound* | *Mass*, g m$^{-3}$ (STPD)	| *Relative Humidity*  	
--- | --- | --- | --- 
P$_4$O$_5$|<2 x 10$^{-3}$ | <0.1
Mg(ClO$_4$)$_2$|<5 x 10$^{-4}$ | <0.1
Mg(ClO$_4$)$_2$&middot;3H$_2$O|<2 x 10$^{-5}$ | <0.1
KOH (fused)|0.002 | <0.1
Al$_2$O$_3$|0.003 | <0.1
H$_2$SO$_4$|0.003 | <0.1
MgO|0.008 | <0.1
NaOH (fused)|0.16 | 0.6
CaBr$_2$|0.2 | 0.8
CaO|0.2 | 0.8
CaCl$_2$|0.14-0.25 | 0.6-1.0
H$_2$SO$_4$ (95%)|0.3 | 1.2
CaCl$_2$ (fused)|0.36 | 1.4
ZnCl$_2$|0.8 | 3.2
ZnBr$_2$|1.1 | 4.4
CuSO$_4$|1.4 | 5.6
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{*}$Relative humidity calculated as 3.98 x mass. 

### Table 10. Water potential (kPa) of NaCl solutions as a function of temperature. From Brown and van Haveren (1972).
<center> <h4>*Temperature*, &deg;C</h4> </center>
```{r table10, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
library(NicheMapR)
knitr::kable(PropAirTable10) # output the table in a format good for HTML/PDF/docx conversion
```

$$
\begin{aligned}
  \dot{x} & = \sigma(y-x) \\
  \dot{y} & = \rho x- y - xz 
\end{aligned}
$$

```{r echo=FALSE, message=FALSE, results='asis', fig.width=7, fig.height=12}
par(mfrow = c(2,1)) # set up for 2 plots in 1 columns
rhs=seq(65,100,5)
plot(rhs,WETAIR(rh=rhs,db=0)$wtrpot*-1,type='l',ylab="WATER POTENTIAL (phi), Pa", xlab="RELATIVE HUMIDITY (%)")
points(rhs,WETAIR(rh=rhs,db=50)$wtrpot*-1,type='l',ylab="WATER POTENTIAL (phi), Pa", xlab="RELATIVE HUMIDITY (%)")
rhs=seq(99.3,100,0.1)
plot(rhs,WETAIR(rh=rhs,db=0)$wtrpot*-1,type='l',ylab="WATER POTENTIAL (phi), Pa", xlab="RELATIVE HUMIDITY (%)")
points(rhs,WETAIR(rh=rhs,db=50)$wtrpot*-1,type='l',ylab="WATER POTENTIAL (phi), Pa", xlab="RELATIVE HUMIDITY (%)")
```

```{r echo=FALSE, message=FALSE, results='asis', fig.width=7, fig.height=6}
tairs=seq(-20,50,5)
plot(tairs,DRYAIR(db=tairs)$emtmax,type='l',ylim=c(0.7E-5,1.2E-5),ylab="WAVELENGTH OF MAXIMUM EMITTANCE FROM BLACK-BODY (lambda_m), m", xlab="AIR TEMPERATURE (t), &deg;C")
```

## References   

Brown RW, Van Havern BP, eds. 1972. Psychrometry in water relations research: proceedings of a symposium on thermocouple psychrometry Utah Agric. Expt. Station.

List RJ. 1971. Smithsonian Meteorological Tables. Smithsonian Institution Press.

Mason EA, Monchick L. 1965. Survey of the equation of state and transport properties of moist gases. Pages 257-272 in 

Wexler A, ed. Humidity and Moisture. Measurement and Control in Science and Industry, vol. 3. New York: Reinhold Publishing Corporation.

Mechtly EA. 1973. The International System of Units: Physical Constants and Conversion Factors. National Aeronautics and Space Administration.

Murray BR. 1976. On the computation of saturation vapor pressure. Journal of Applied Meteorology 6:203-204.

Symbols Committee of the Royal Society. 1975. Quantities, Units and Symbols. The Royal Society.


