---
title: "Endotherm Model Tutorial"
author: "Michael Kearney"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Endotherm Model Tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 eval = TRUE, tidy.opts=list(width.cutoff=60), tidy=TRUE  
)
```

## Overview

This document shows how to use the `endoR.R` function of NicheMapR to compute endotherm heat and mass balances. This function collates all the different steps in [Endotherm Components Tutorial](endotherm-components-tutorial.html) into a single function that can be applied to vectors of data.

## Single calculation example

Here is how you run the calculation with all default values except specifying a local air temperature of 0 deg C.

```{r}
library(NicheMapR)
library(knitr)
endo.out <- endoR(TA = 0)
kable(endo.out[, 1:9])
kable(endo.out[, 10:19])
kable(endo.out[, 20:29])
kable(endo.out[, 30:39])
kable(endo.out[, 40:49])
kable(endo.out[, 50:60])
```

## Metabolic chamber example

Now try a sequence of air temperatures for a specified core temperature, mass, fur/feather depth and base skin wetness, with the option to sweat at 0.25 % increments to cool down if needed. Note that the default settings make the environment a blackbody one (all surface and air temperatures the same) with low (5%) relative humidity and low wind speed (0.1 m/s), as is often the case in metabolic chamber experiments.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
library(NicheMapR)
# environment
TAs <- seq(0, 50, 1) # air temperature (deg C)
VEL <- 0.002 # wind speed (m/s)
vd <- WETAIR(rh = 30, db = 40)$vd # Weather and Schoenbaechler had 16.7 mm Hg above 40 deg C = 30% RH at 40 deg C
vd_sat <- WETAIR(rh = 100, db = TAs)$vd # Weather and Schoenbaechler had 16.7 mm Hg above 40 deg C = 30% RH at 40 deg C
exp_rh <- vd / vd_sat * 100
exp_rh[exp_rh > 100] <- 100
exp_rh[TAs < 30] <- 15
hum <- exp_rh#rep(humidity,96)

# core temperature
TC <- 38 # core temperature (deg C)
TCMAX <- 43 # maximum core temperature (deg C)
RAISETC <- 0.25 # increment by which TC is elevated (deg C)

# size and shape
AMASS <- 0.0337 # mass (kg)
GMREF <- 2.5 # start off near to a sphere (-)
GMULTMAX <- 2.5 # maximum ratio of length to width/depth
UNCURL <- 0.1 # allows the animal to uncurl to GMULTMAX, the value being the increment GMULT is increased per iteration
BIRD <- 1 # use bird surface area relations

# feather properties
DHAIRD <- 30E-06 # hair diameter, dorsal (m)
DHAIRV <- 30E-06 # hair diameter, ventral (m)
LHAIRD <- 23.1E-03 # hair length, dorsal (m)
LHAIRV <- 22.7E-03 # hair length, ventral (m)
ZFURD <- 5.8E-03 # fur depth, dorsal (m)
ZFURV <- 5.6E-03 # fur depth, ventral (m)
RHOD <- 8000E+04 # hair density, dorsal (1/m2)
RHOV <- 8000E+04 # hair density, ventral (1/m2)
REFLD <- 0.248  # fur reflectivity dorsal (fractional, 0-1)
REFLV <- 0.351  # fur reflectivity ventral (fractional, 0-1)

# physiological responses
SKINW <- 0.1 # base skin wetness (%)
MXWET <- 0.1 # maximum skin wetness (%)
SWEAT <- 0.25 # intervals by which skin wetness is increased (%)
Q10s <- rep(1, length(TAs))
Q10s[TAs >= TCMAX] <- 2 # assuming Q10 effect kicks in only after air temp rises above TCMAX
QBASAL <- 10 ^ (-1.461 + 0.669 * log10(AMASS * 1000)) # basal heat generation (W)
DELTAR <- 5 # offset between air temeprature and breath (°C)
EXTREF <- 15 # O2 extraction efficiency (%)
PANTING <- 0.1 # turns on panting, the value being the increment by which the panting multiplier is increased up to the maximum value, PANTMAX
PANTMAX <- 10# maximum panting rate - multiplier on air flow through the lungs above that determined by metabolic rate

ptm <- proc.time() # start timing
endo.out <- lapply(1:length(TAs), function(x){endoR(TA = TAs[x], VEL = VEL, TC = TC, TCMAX = TCMAX, RH = hum[x], AMASS = AMASS, GMREF = GMREF, GMULTMAX = GMULTMAX, SKINW = SKINW, SWEAT = SWEAT, MXWET = MXWET, Q10 = Q10s[x], QBASAL = QBASAL, DELTAR = DELTAR, DHAIRD = DHAIRD, DHAIRV = DHAIRV, LHAIRD = LHAIRD, LHAIRV = LHAIRV, ZFURD = ZFURD, ZFURV = ZFURV, RHOD = RHOD, RHOV = RHOV, REFLD = REFLD, RAISETC = RAISETC, PANTING = PANTING, PANTMAX = PANTMAX, EXTREF = EXTREF, UNCURL = UNCURL, BIRD = BIRD)}) # run endoR across environments
proc.time() - ptm # stop timing

endo.out <- do.call("rbind", lapply(endo.out, data.frame)) # turn results into data frame

QGEN <- endo.out$RESPGEN # metabolic rate (W)
H2O <- endo.out$EVAP.G.H # g/h water evaporated
TFA_D <- endo.out$TFA_D # dorsal fur surface temperature
TFA_V <- endo.out$TFA_V # ventral fur surface temperature
TskinD <- endo.out$TSKIN_D # dorsal skin temperature
TskinV <- endo.out$TSKIN_V # ventral skin temperature
TCs <- endo.out$TC # core temperature

par(mfrow = c(2, 2))
par(oma = c(2, 1, 2, 2) + 0.1)
par(mar = c(3, 3, 1.5, 1) + 0.1)
par(mgp = c(2, 1, 0))
plot(QGEN ~ TAs, type = 'l', ylab = 'metabolic rate, W', xlab = 'air temperature, deg C', ylim = c(0.2, 1.2))
points(Weathers1976Fig1$Tair, Weathers1976Fig1$mlO2gh * 20.1 / 3600 * (AMASS * 1000), pch = 16, col = 2)
plot(H2O ~ TAs, type = 'l', ylab = 'water loss, g/h', xlab = 'air temperature, deg C', ylim = c(0, 1.5))
points(Weathers1976Fig3$Tair, Weathers1976Fig3$mgH2Ogh * AMASS, pch = 16, col = 2)
plot(TFA_D ~ TAs, type = 'l', col = 'grey', ylab = 'fur, skin and core temperature, deg C', xlab = 'air temperature, deg C', ylim = c(10, 50))
points(TFA_V ~ TAs, type = 'l', col = 'grey', lty = 2)
points(TskinD ~ TAs, type = 'l', col = 'orange')
points(TskinV ~ TAs, type = 'l', col = 'orange', lty = 2)
points(TCs ~ TAs, type = 'l', col = 'red')
points(Weathers1976Fig2$Tair, Weathers1976Fig2$Tb, pch = 16, col = 2)
plot(endo.out$AIRVOL * 1e6 / 60 ~ TAs, ylim=c(0,250),  lty = 1, xlim=c(-5,50), main = "minute volume", ylab = "ml / min", xlab=paste("air temperature (deg C)"), type = 'l') 
points(Weathers1976Fig5$breaths_min * (13.2 * AMASS ^ 1.08) * ((Weathers1976Fig5$Tair + 273.15) / 273.15) ~ Weathers1976Fig5$Tair, col='red', pch = 16) # tidal volume allometry from Lasiewski, R. C., and W. A. Calder. 1971, correcting volume according to PV = nRT equation, where V_2 = T_2 * V_1 / T_2, and T_1 is at STP, so 0 deg C
```

## Microclimate model integration example

In this final example, the microclimate model is run for a location in central Australia and the then the relevant outputs are sent into the endotherm model along with the same animal parameters as above.

```{r, fig.width=7, fig.height=5, fig.show = "hold", message=FALSE, warnings=FALSE}
library(NicheMapR)
loc <- c(131.05, -22.75)
Usrhyt <- 0.07
maxshade <- 100
micro <- micro_global(loc = loc, Usrhyt = Usrhyt, maxshade = maxshade)

metout <- as.data.frame(micro$metout)
soil <- as.data.frame(micro$soil)
shadmet <- as.data.frame(micro$shadmet)
shadsoil <- as.data.frame(micro$shadsoil)
days<-rep(seq(1,12),24)
days<-days[order(days)]
dates<-days+metout$TIME/60/24-1 # dates for hourly output

warm <- 0
TAs <- metout$TALOC
TAREFs <- metout$TAREF
TSKYs <- metout$TSKYC
TGRDs <- soil$D0cm
VELs <- shadmet$VLOC
RHs <- metout$RHLOC
QSOLRs <- metout$SOLR 
Zs <- shadmet$ZEN
ELEV <- micro$elev
ABSSB <- 1-micro$REFL

# core temperature
TC <- 38 # core temperature (deg C)
TCMAX <- 43 # maximum core temperature (deg C)
RAISETC <- 0.25 # increment by which TC is elevated (deg C)

# size and shape
AMASS <- 0.0337 # mass (kg)
GMREF <- 1.1 # start off near to a sphere (-)
GMULTMAX <- 5 # maximum ratio of length to width/depth
UNCURL <- 0.1 # allows the animal to uncurl to GMULTMAX, the value being the increment GMULT is increased per iteration
BIRD <- 1

# feather properties
DHAIRD = 30E-06 # hair diameter, dorsal (m)
DHAIRV = 30E-06 # hair diameter, ventral (m)
LHAIRD = 23.1E-03 # hair length, dorsal (m)
LHAIRV = 22.7E-03 # hair length, ventral (m)
ZFURD = 5.8E-03 # fur depth, dorsal (m)
ZFURV = 5.6E-03 # fur depth, ventral (m)
RHOD = 8000E+04 # hair density, dorsal (1/m2)
RHOV = 8000E+04 # hair density, ventral (1/m2)
REFLD = 0.248  # fur reflectivity dorsal (fractional, 0-1)
REFLV = 0.351  # fur reflectivity ventral (fractional, 0-1)

# physiological responses
SKINW <- 0.1 # base skin wetness (%)
MXWET <- 0.1 # maximum skin wetness (%)
SWEAT <- 0.25 # intervals by which skin wetness is increased (%)
Q10 <- 1 # Q10 effect of body temperature on metabolic rate (-)
QBASAL <- 10 ^ (-1.461 + 0.669 * log10(AMASS * 1000)) # basal heat generation (W)
DELTAR <- 5 # offset between air temeprature and breath (°C)
EXTREF <- 15 # O2 extraction efficiency (%)
PANTING <- 0.1 # turns on panting, the value being the increment by which the panting multiplier is increased up to the maximum value, PANTMAX
PANTMAX <- 10# maximum panting rate - multiplier on air flow through the lungs above that determined by metabolic rate

ptm <- proc.time() # start timing
endo.out <- lapply(1:length(TAs), function(x){endoR(TA = TAs[x], TAREF = TAREFs[x], TSKY = TSKYs[x], TGRD = TGRDs[x], VEL = VELs[x], RH = RHs[x], QSOLR = QSOLRs[x], Z = Zs[x], ELEV = ELEV, ABSSB = ABSSB, TC = TC, TCMAX = TCMAX, AMASS = AMASS, GMREF = GMREF, GMULTMAX = GMULTMAX, SKINW = SKINW, SWEAT = SWEAT, Q10 = Q10, QBASAL = QBASAL, DELTAR = DELTAR, DHAIRD = DHAIRD, DHAIRV = DHAIRV, LHAIRD = LHAIRD, LHAIRV = LHAIRV, ZFURD = ZFURD, ZFURV = ZFURV, RHOD = RHOD, RHOV = RHOV, REFLD = REFLD, RAISETC = RAISETC, PANTING = PANTING, PANTMAX = PANTMAX, EXTREF = EXTREF, UNCURL = UNCURL, BIRD = BIRD, SHADE = 0)})
proc.time() - ptm
endo.out <- do.call("rbind", lapply(endo.out, data.frame))

QGEN <- endo.out$RESPGEN # metabolic rate (W)
H2O <- endo.out$EVAP.G.H # g/h water evaporated
TFA_D <- endo.out$TFA_D # dorsal fur surface temperature
TFA_V <- endo.out$TFA_V # ventral fur surface temperature
TskinD <- endo.out$TSKIN_D # dorsal skin temperature
TskinV <- endo.out$TSKIN_V # ventral skin temperature
TCs <- endo.out$TC # core temperature
SkinW <- endo.out$SKINW # skin wetness (%)
Pant <- endo.out$PANT # panting multiplier (-)

par(mfrow = c(2, 2))
par(oma = c(2, 1, 2, 2) + 0.1)
par(mar = c(3, 3, 1.5, 1) + 0.1)
par(mgp = c(2, 1, 0))
plot(QGEN ~ dates, type = 'l', ylab = 'metabolic rate, W', xlab = 'time', ylim = c(0, QBASAL * 5))
plot(H2O ~ dates, type = 'l', ylab = 'water loss, g / h', xlab = 'time', ylim = c(0, 2))
plot(TFA_D ~ dates, type = 'l', col = 'grey', ylab = 'fur, skin and core temperature, deg C', xlab = 'time', ylim = c(0, 60))
points(TFA_V ~ dates, type = 'l', col = 'grey', lty = 2)
points(TskinD ~ dates, type = 'l', col = 'orange')
points(TskinV ~ dates, type = 'l', col = 'orange', lty = 2)
points(TCs ~ dates, type = 'l', col = 'red')
plot((H2O / (AMASS * 1000)) * 1000 ~ dates, type = 'l', ylab = 'water loss, % body mass / h', xlab = 'time', ylim = c(0, 50))
```
