---
title: 'Microclimate Model Input Data'
author: Michael R. Kearney
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microclimate Model Input Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
 eval = TRUE
)
```


## Model mode settings
```{r table1, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
writecsv                  | -                 | 0 (off) or 1 (on) | make Fortran program write output as csv files  
microdaily                | -                 | 0 (off) or 1 (on) | run in daily mode (initial conditions from previous day)
runshade                  | -                 | 0 (off) or 1 (on) | run the model twice, once for each shade level
runmoist                  | -                 | 0 (off) or 1 (on) | run soil moisture model
snowmodel                 | -                 | 0 (off) or 1 (on) | run the snow model
hourly                    | -                 | 0 (off) or 1 (on) | run the model from hourly weather inputs
IR                        | -                 | 0 or 1            | longwave radiation algorithm
message                   | -                 | 0 or 1            | integrator messages
fail                      | -                 | integer            | integrator failure count before quitting
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

## Time and location parameters
```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
doynum                    | days              | positive integer  | number of days to run the model
doy                       | day-of-year       | 1-365             | vector of days of year (length must equal doynum)
idayst                    | -                 | 1-doynum          | start day (usually 1)
ida                       | -                 | 1-doynum          | end day (usually value of doynum)
HEMIS                     | -                 | 1 (N) or 2 (S)    | hemisphere to run
ALAT                      | degrees           | 0-90              | latitude (degrees)
AMINUT                    | dec. minutes      | 0-60              | latitude (minutes)
ALONG                     | degrees           | 0-180             | longitude (degrees)
ALMINT                    | dec. minutes      | 0-60              | longitude (minutes)
ALREF                     | degrees           | 0-180             | reference longitude (degrees) for time zone
EC                        | -                 | 0.0034 to 0.058   | eccenricity of the earth's orbit (presently 0.0167238)
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

## Air and wind vertical profile parameters
```{r table3, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
RUF                       | m                 | 0.01-200          | roughness height
Refhyt                    | m                 | 0.50-10           | reference height for air temp, wind speed and humidity input data
Usrhyt                    | m                 | > 0.005, < Refhyt | local height at which to compute air temp, wind speed and humidity 
ZH                       | m                 | > 0 (or else not used)   | heat transfer roughness height (m)
D0                       | m                 | > 0                      | zero plane displacement correction factor (m) used with ZH
Z01                       | m                 | 0 or (> Z02, < ZH2)   | 1st segment, roughness height$^1$
Z02                       | m                 | 0 or (> RUF, < Z01)   | 2nd segment, roughness height$^1$
ZH1                       | m                 | 0 or (> ZH2, < Refhyt)| 1st segment, height above surface$^1$
ZH2                       | m                 | 0 or (> RUF, < ZH1)   | 2nd segment, height above surface$^1$
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^1$Set to 0 if no experimental data

## Radiation-related parameters
```{r table4, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
SLES                      | -                 | 0-1               | substrate longwave IR emissivity$^1$ 
REFLS                      | -                 | 0-1               | substrate solar reflectivity$^1$ 
CMH2O                     | cm                | 0.1-2             | precipitable cm H$_2$O in air column
TAI                       | -                 | 1-doynum          | aerosol optical extinction coefficients$^{1,2}$ 
solonly                   | -                | 0-1             | flag to do only solar calcs
lamb                     | -                | 0-1             | flag to return wavelength-specific solar radiation output
IUV                      | -                | 0-1             | flag to use gamma function for scattered solar radiation$^{3}$
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Vector of length = doynum

$^{2}$Vector of length 111 for wavelengths between 290 and 4000 nm. Wavelenghth increments are: 290,295,300,305,310,315,320,330,340,350,360,370,380,390,400,420,440,460,480,500,
520,540,560,580,600,620,640,660,680,700,720,740,760,780,800,820,840,860,880,900,
920,940,960,980,1000,1020,1080,1100,1120,1140,1160,1180,1200,1220,1240,1260,1280,
1300,1320,1380,1400,1420,1440,1460,1480,1500,1540,1580,1600,1620,1640,1660,1700,
1720,1780,1800,1860,1900,1950,2000,2020,2050,2100,2120,2150,2200,2260,2300,2320,
2350,2380,2400,2420,2450,2490,2500,2600,2700,2800,2900,3000,3100,3200,3300,3400,
3500,3600,3700,3800,3900,4000

$^{3}$Computationally intensive

## Terrain and shading parameters
```{r table5, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
ALTT                      | m                 | 0-10,000          | elevation
slope                     | decimal degrees   | 0-90              | slope
azmuth                    | decimal degrees   | 0-360             | aspect (0 is north)
hori                      | decimal degrees   | 0-90              | horizon angles$^1$  
VIEWF                     | -                 | 0-1               | view factor to sky$^2$ 
MINSHADES                 | %                 | 0-100             | minimum shade$^3$  
MAXSHADES                 | %                 | 0-100             | maximum shade$^{3,4}$ 
PCTWETS                   | %                 | 0-100             | percentage of substrate unit area that is wet$^{3,5}$  
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$angle to the effective horizon, due to e.g. hills, buildings, in 24 directions from 0 degrees azimuth (north) clockwise in 15 degree intervals

$^{2}$Fraction of sky obscured by terrain

$^{3}$vector of length = doynum

$^{4}$must be greater than minimum shade

$^{5}$drives evaporation - set internally when runmoist==1

## Substrate profile settings
```{r table6, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
DEP                       | cm                | 0-1000            | vector of 10 depths$^1$
ERR                       | -                 | >0                | integrator error (typically 1.5-2)
tannul                    | &deg;C            | -80 - +60         | annual mean temperature
soilinit                  | &deg;C            | -80 - +60         | initial substrate temperature profile$^1$
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Acting as nodes for substrate heat budget calculations. Must start at 0 and nodes must be closely spaced near the surface. Typical depth profile c(0., 2.5,  5.,  10.,  15,  20,  30,  50,  100,  200)

$^{2}$Vector of length 10, corresponding to depths specified in variable DEP

## Time varying environmental data
```{r table7, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *          |	*Allowed Range*   |	*Description*
------------------------- | ----------------- | ----------------- | --------------
TIMINS                    | h                 | 0-23              | time of minima for air temp, wind, humidity and cloud cover$^1$
TIMAXS                    | h                 | 0-23              | time of maxima for air temp, wind, humidity and cloud cover$^2$
TMINN                     | &deg;C            | -80 - +60         | minimum air temperature (at reference height, Refhyt)$^3$
TMAXX                     | &deg;C            | -80 - +60         | maximum air temperature (at reference height, Refhyt)$^3$
RHMINN                    | %                 | 0-100             | minimum relative humidity (at reference height, Refhyt)$^3$
RHMAXX                    | %                 | 0-100             | maximum relative humidity (at reference height, Refhyt)$^3$
WNMINN                    | m s$^{-1}$        | 0-100             | minimum wind speed (at reference height, Refhyt)$^3$
WNMAXX                    | m s$^{-1}$        | 0-100             | maximum wind speed (at reference height, Refhyt)$^3$
CCMINN                    | %                 | 0-100             | minimum cloud cover; vector of length = doynum
CCMAXX                    | %                 | 0-100             | maximum cloud cover; vector of length = doynum
RAINFALL                  | mm                | 0-2000            | daily total rainfall; vector of length = doynum
tannulrun                 | &deg;C            | -80 - +60         | daily deep soil temperature; vector of length = doynum
moists                    | decimal %         | 0-1               | predefined soil daily moisture profile through time$^4$
TAIRhr                    | &deg;C            | -80 - +60         | hourly air temperature (at reference height, Refhyt)$^5$
RHhr                      | %                 | 0-100             | hourly relative humidity (at reference height, Refhyt)$^5$
WNhr                      | m s$^{-1}$        | 0-100             | hourly wind speed (at reference height, Refhyt)$^5$
CLDhr                     | %                 | 0-100             | hourly cloud cover$^5$
SOLRhr                    | W m$^{2}$         | 0-1367            | hourly solar radiation$^5$
RAINhr                    | mm                | 0-2000            | hourly rainfall$^5$
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Vector of 4 integers, air temp & wind mins relative to sunrise, humidity and cloud cover mins relative to solar noon - typical TIMINS vector c(0, 0, 1, 1).  

$^{2}$Vector of 4 integers, air temp & wind maxs relative to solar noon, humidity and cloud cover maxs relative to sunrise - typical TIMAXS vector c(1, 1, 0, 0).

$^{3}$Vector of length = doynum.

$^{4}$Matrix of 10 rows (depths) and doynum columns. The first column is used to specify the initial soil moisture when running the soil moisture model.

$^{5}$Vector of length = doynum*24.

## Substrate properties
```{r table8, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *             |	*Allowed Range*   |	*Description*
------------------------- | -------------------- | ------------------ | --------------
Numtyps                   | -                    | 1-10               | number of substrate types
Nodes$^{1,2}$             | -                    | 1-10               | nodes where substrate type transitions occur
soilprops[,1]$^{3}$       | Mg m$^{-3}$          | >0                 | bulk density (must not exceed mineral density)
soilprops[,2]$^{3}$       | m$^{3}$ m$^{-3}$     | >0                 | volumetric water content at saturation$^{4}$ 
soilprops[,3]$^{3}$       | W m$^{-1}$ K$^{-1}$  | >0                 | thermal conductivity
soilprops[,4]$^{3}$       | J kg$^{-1}$ K$^{-1}$ | >0                 | specific heat capacity
soilprops[,5]$^{3}$       | Mg m$^{-3}$          | >0                 | mineral density
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$Matrix of 10 rows (depths) and doynum columns

$^{2}$The number of nodes specified should correspond with the variable numtypes, and in turn with the number of substrate types specified in soilprops in the soil props matrix. E.g. for a uniform profile, the first row of Nodes would be 10 (deepest soil node - the subsequent nodes all being left as 0), Numtyps would be 1, and only one and only the first row of soilprops would be filled, the rest being zero. At the other extreme, if all nodes were to have a different substrate type, values for a given row of each Nodes column would be 1, 2, ..., 10 and a different value would be specified for each row of each soilprops column.  

$^{3}$Matrix of 10 rows (depths) and 6 columns

$^{4}$at a matric potential of 0.1 bar

## Soil moisture parameters
```{r table9, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *             |	*Allowed Range*  |	*Description*
------------------------- | -------------------- | ----------------- | --------------
PE                        | J kg$^{-1}$          | 0.7-3.7           | air entry water potential; vector of 19 values$^{1}$
KS                        | kg s m$^{-3}$        | 1 x $10^{-5}$     | saturated conductivity; vector of 19 values$^{1}$
BB                        | -                    | 1.7-7.6           | Campbell's 'b' parameter; vector of 19 values$^{1}$
BD                        | Mg m$^{-3}$          | >0                | bulk density (should be matched to same variable in the soilprops matrix)
DD                        | Mg m$^{-3}$          | >0                | soil mineral density (should be matched to same variable in the soilprops matrix)
L                         | m$^{3}$ m$^{-3}$     | 0-90              | root density; vector of 19 values$^{1}$  
R1                        | m                    | 0-90              | root radius  
RW                        | m$^{3}$ kg$^{-1}$ s$^{-1}$     | 0-90              | resistance per unit length of root  
RL                        | m$^{3}$ kg$^{-1}$ s$^{-1}$     | 0-90              | resistance per unit length of leaf  
PC                        | J kg$^{-1}$          | 0-90              | critical leaf water potential for stomatal closure  
SP                        | -                    | >0                | stability parameter for stomatal closure equation
IM                        | kg                   | >0                | maximum allowable mass balance error
MAXCOUNT                  | -                    | >0                | maximum iterations for mass balance
LAI                       | -                    | 0-90              | leaf area index, used to partition traspiration/evaporation from PET 
rainmult                  | -                    | >0                | rainfall multiplier to impose a catchment 
maxpool                   | mm                   | >0                | max depth for surface water pooling$^{2}$  
evenrain                  | -                    | 1 or 2            | even rainfall over 24hrs (1) or one event at midnight (2)
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$The 19 values represent the 10 depths as specified in the DEP vector, and an extra 9 depths in between. 

$^{2}$To account for runoff - can be used to simulate a wetland.

## Snow model parameters
```{r table10, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *             |	*Allowed Range*   |	*Description*
------------------------- | -------------------- | ------------------ | --------------
snowtemp                  | &deg;C               | -80 - 80           | temperature at which precipitation falls as snow
snowdens                  | Mg m$^{-3}$          | 0.05-1             | snow density
densfun                   | -                    | -                  | parameters of snow density function$^{1}$ 
snowmelt                  | decimal %            | 0-1                | proportion of calculated snowmelt that doesn't refreeze  
undercatch                | -                    | >=1                | undercatch multipier for converting rainfall to snow 
rainmelt                  | -                    | 0-2                | parameter for rain melting snow$^{2}$ 
snowcond                  | -                    | >0                 | effective thermal conductivity of snow $W/mK$ 
intercept                 | -                    | 0-1                | proportion of snow intercepted by vegetation under shaded conditions 
grasshade                 | -                    | 0 or 1             | if 1, shade is not present when snow is present (because shade is cast by grass/low veg) 
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```

$^{1}$ Four values, c(a, b, c, d) representing either 1) slope (a) and intercept (b, Mg m-3) of model of snow density as a linear function of snowpack age (days) if first two values are nonzero, or, if all values are non-zero, b) following the exponential function of Sturm et al. 2010 J. of Hydromet. 11:1380-1394 where density = (a - b) x (1 - exp(-c x depth - d x age)) + b, where a and b are the maximum and minimum densities, respectively, in Mg m-3, and c and d are fitted parameters. If all values are zero, i.e. c(0,0,0,0), then a fixed density is used as specified by parameter snowdens


$^{2}$For equation from Anderson's SNOW-17 model that melts snow with rainfall as a function of air temp (typical value 1.25)

## Intertidal simulation parameters (currently experimental and untested)
```{r table11, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- "
*Name*                    | *Units *             |	*Allowed Range*   |	*Description*
------------------------- | -------------------- | ------------------ | --------------
tides[1:doynum,1]         | -                    | 0 or 1             | tide in (1) or out(0); vector of length doynum
tides[1:doynum,2]         | &deg;C               | -80 - 80           | sea water temperature; vector of length doynum
tides[1:doynum,3]         | %                    | -                  | % surface wetness from wave splash; vector of length doynum
"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```
